---
title: "HW 3 Solutions"
author: "Tom Fletcher"
date: "October 19, 2017"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```


__This solution set is for use by students in CS 3130 / ECE 3530 in Fall 2016 ONLY. Distribution outside of the class will be considered cheating.__

# Problem 1

We just enumerate all of the events defined by the random variable $X$:

$$\{X = 1\} = \{HTH, THT\}$$

$$\{X = 2\} = \{HHT, HTT, THH, TTH\}$$

$$\{X = 3\} = \{HHH, TTT\}$$

Because each outcome is equally likely, with probability $\frac{1}{8}, we get the pmf:

$$p_X(1) = \frac{1}{4}, \quad p_X(2) = \frac{1}{2}, \quad p_X(3) = \frac{1}{4}.$$

# Problem 2

Set up the sample space:
```{r}
Omega = c('HHH', 'HHT', 'HTH', 'HTT', 'THH', 'THT', 'TTH', 'TTT')
```

Run 10,000 simulations:
```{r}
numSims = 10000
sims = sample(Omega, size = numSims, replace = TRUE)
```

Estimate the pmf for $X$ by counting the occurrences for each value:
```{r}
p = numeric(3)
p[1] = sum(sims == 'HTH' | sims == 'THT') / numSims
p[2] = sum(sims == 'HHT' | sims == 'HTT' | sims == 'THH' | sims == 'TTH') / numSims
p[3] = sum(sims == 'HHH' | sims == 'TTT') / numSims
```

Finally, the simulated pmf for $X$ is

$$p_x \approx `r p`$$

# Problem 3

We first identify this as a __geometric distribution__ because we will keep playing the lotteries until our first success (we win at least one). Now, we need to define $p$, the probability of success, that is, the probability of winning at least one lottery. The probability of winning both lotteries is given by $p_1 \times p_2$ because they are independent. So, using inclusion exclusion, the probability of winning at least one is
$$p = p_1 + p_2 - p_1 \times p_2.$$


# Problem 4

In the following answers, the random variable $X$ will denote the number
of errors. An error has probability 10\%. You could also ``flip'' this and work
with the number of successes: $Y = 50 - X$. The probability of a success is
90\%.

## (a)

(1) Binomial pmf with $n = 50,$ $p = 0.1,$ $k = 8$ ($k$ is number of
errors) or $n = 50,$ $p = 0.9,$ $k = 42$ ($k$ is number of successes). Formula
is $P(X = 8) = {50 \choose 8}\, 0.1^8\, 0.9^{42}$.

(2) 
```{r}
dbinom(8, 50, 0.1)
```
or 

```{r}
dbinom(42, 50, 0.9)
```

## (b)
(1) Binomial cdf with $n = 50,$ $p = 0.1,$ $k = 5$ or $n = 50,$ $p =
0.9,$ $k = 45$. Formula is $P(X \leq 5) = \sum_{i=1}^{5} {50 \choose i} \,
0.1^i \, 0.9^{50 - i}.$

(2) 
```{r}
pbinom(5, 50, 0.1)
```
or 

```{r}
1 - pbinom(44, 50, 0.9)
```

## (c)
(1) Binomial pmf with $n = 50,$ $p = 0.1,$ $k = 0$ or $n = 50,$ $p =
0.9,$ $k = 50$. Formula simplifies to $P(X = 0) = 0.9^{50}$

(2) 
```{r}
dbinom(0, 50, 0.1)
```
or simply 
```{r}
0.9^50
```


# Problem 5

Simulate the number of successes for the system using binomial random variable (`rbinom`):
```{r}
N = 10000
x = rbinom(N, size = 50, prob = 0.9)
```

## (a)
Simulated probability of exactly 8 errors:
```{r}
sum(x == 42) / N
```

## (b)
Simulated probability of 5 or fewer errors:
```{r}
sum(x >= 45) / N
```

## (c)
Simulated probability of a perfect day:
```{r}
sum(x == 50) / N
```

# Problem 5

## (a)

More girls: Let $X$ be a geometric random variable (with $p =
0.5$). Then ``more girls'' happens if $X = 1$, ``boys = girls'' happens if $X =
2$, and ``more boys'' happens if $X > 2$. $P(X = 1) = 0.5$, $P(X = 2) = 0.25$,
so $P(X > 2) = 0.25$.

## (b)

Formula for $k$ boys:

$P(X = k+1) = 0.5^{k + 1}$

## (c)

Still more girls: $P(X = 1) = 0.4$ is still higher than $P(X > 2) =
0.36$.

## (d)

Formula for $k$ boys: 

$P(X = k+1) = 0.6^k \cdot 0.4$
