---
title: "HW 4 Solutions"
author: "Tom Fletcher"
date: "October 19, 2017"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```


__This solution set is for use by students in CS 3130 / ECE 3530 in Fall 2017 ONLY. Distribution outside of the class will be considered cheating.__

# Problem 1

Throughout this problem, we are using $X \sim N(85, 5^2)$, that is,
a normal distribution with parameters $\mu = 85$ and $\sigma = 5$.

## (a)

Using the cdf and the complement rule, the probability of being greater than 90 is $P(X > 90) = 1 - P(X \leq 90)$:

```{r}
1 - pnorm(90, mean = 85, sd = 5)
```

## (b)

Quantile at 0.95:

```{r}
qnorm(0.95, mean = 85, sd = 5)
```

## (c)

Using the cdf twice, we get $P(70 \leq X \leq 80) = P(X \leq 80) - P(X \leq 70)$:

```{r}
pnorm(80, mean = 85, sd = 5) - pnorm(70, mean = 85, sd = 5)
```

# Problem 2

First, simple integration gives the cdf:
$$F(x) =
\begin{cases}
0 & \text{for $x < 0$}\\
2x^2 & \text{for $0 \leq x \leq \frac{1}{2}$}\\
4x - 2x^2 - 1 & \text{for $\frac{1}{2} \leq x \leq 1$}\\
1 & \text{for $x > 1$}\\
\end{cases}$$

## (a)
$F(0.55) = 4 \cdot 0.55 - 2 \cdot 0.55^2 - 1 = 0.595$

## (b)
Find the 50th percentile by solving for the $x$ that satisfies:
$F(x) = 0.5$

This occurs at $x = 0.5$.


# Problem 3

Let $X$ be the time between user requests. We know that $X \sim
\mathit{Exp}(\lambda)$, but we don't know $\lambda$. The event "the next user has to wait" is $\{X \leq 1\}$ because if the next request comes in under one minute, we will still be processing the previous request. We want to know the rate $\lambda$ that will cause this event to have 10\% probability. So, using the exponential cdf, we get $0.10 = P(X \leq 1) = 1 - \exp(-\lambda)$. Solving for $\lambda$ gives $\lambda = -\log(0.9) \approx `r -log(0.9)`$ (in requests/min). Also, any slower rate will result in an even lower probability of waiting.


# Problem 4

We are looking for the $0.5$ quantile, $q_{0.5}$. (Let's call the median
$m = q_{0.5}$ to simplify the notation.) This will be $m = F^{-1}(0.5)$, where
$F$ is the Pareto cdf. In other words, we can write $0.5 = F(m)$, and solve for
the unknown $m$. Writing this out, we have $0.5 = F(m) = 1 -
\frac{1}{m^\alpha}$. Solving for $m$, we get $m = 2^{\frac{1}{\alpha}}$, i.e.,
$m = \sqrt[\alpha]{2}$
