---
title: "HW 4 Solutions"
author: "Tom Fletcher"
date: "November 13, 2017"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```


__This solution set is for use by students in CS 3130 / ECE 3530 in Fall 2017 ONLY. Distribution outside of the class will be considered cheating.__

# Problem 1

## (a)

Net winnings are $-1$, $0$, or $1$ jellybean after paying the 1 jellybean. This gives:

$$E[X] = -1 \times \frac{2}{6} + 0 \times \frac{2}{6} + 1 \times \frac{2}{6} = 0$$

Or, if you are using winnings before the -1 jellybean, you would get:
$$E[X] = 0 \times \frac{2}{6} + 1 \times \frac{2}{6} + 2 \times \frac{1}{6} = 1$$

## (b)

$$E[X^2] = (-1)^2 \times \frac{2}{6} + 0^2 \times \frac{2}{6} + 1^2 \times \frac{2}{6} = \frac{2}{3}$$

$$\mathrm{Var}(X) = E[X^2] - E[X]^2 = \frac{2}{3}$$

The variance is the same whether you subtract the one jellybean or not!

## (c)

It doesn't matter if you play or not! In the long run, you end up with the same amount of jellybeans as you started.


# Problem 2

```{r}
x = sample(c(-1, 0, 1), 100000, prob = c(1/3, 1/3, 1/3), replace = TRUE)
mean(x)
```

# Problem 3

Define three random variables:

* $X_1$ = "number of rolls to get 1, 2, or 3"
* $X_2$ = "number more rolls it takes to get next number I'm looking for"
* $X_3$ = "number more rolls it takes to get final number"

These random variables are geometric random variables with probability $p = \frac{1}{2}, \frac{1}{3}, \frac{1}{6}$, respectively.

Then the total number of rolls it takes to get all three (1,2,3) is $X = X_1 + X_2 + X3$. The expected value of $X$, using linearity of expectation and that geometric expectation is $\frac{1}{p}, is

$$E[X] = E[X_1 + X_2 + X_3] = E[X_1] + E[X_2] + E[X_3] = \frac{1}{1/2} + \frac{1}{1/3} + \frac{1}{1/6} = 11$$

# Problem 4

## (a)
$$1 = F(c) = \int_0^c 2 - 2x dx = 2 c - c^2$$
Solving for $c$ in quadratic formula gives: $c = 1$.

## (b)
$$E[X] = \int_0^1 x (2 - 2x) dx = x^2 - \frac{2}{3}x^3 \Big|_{x=0}^{x=1} = 1 - \frac{2}{3} = \frac{1}{3}$$

## (c)
$$E[X^2] = \int_0^{0.5} x^2 (2 - 2x) dx = \frac{2}{3} x^3 - \frac{1}{2} x^4 \Big|_{x=0}^{x=1} = \frac{2}{3} - \frac{1}{2} = \frac{1}{6}$$

$$\mathrm{Var}(X) = E[X^2] - E[X]^2 = \frac{1}{6} - \frac{1}{3}^2 = \frac{1}{18}$$

## (d)

$$E[3X + 2] = 3 E[X] + 2 = 3$$

## (e)

$$\mathrm{Var}(3X + 2) = 9 \mathrm{Var}(X) = \frac{1}{2} $$

# Problem 5

$$E[\cos(X)] = \int_{0}^\pi \cos(x) \frac{1}{\pi} dx = \frac{\sin(x)}{\pi} \Big|_{x=0}^{x=\pi} = 0$$

# Problem 6

```{r}
mean(cos(runif(100000, 0, pi)))
```

A quick check shows this is close to the true value 0.
