---
title: "HW 8 Solutions"
author: "Tom Fletcher"
date: "December 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

__This solution set is for use by students in CS 3130 / ECE 3530 in Fall 2017 ONLY. Distribution outside of the class will be considered cheating.__

## Problem 1

Read in the cardiac data:
```{r}
cardiac = read.csv("cardiac.dat")
```

Here is the contingency table. Male/Female are rows, HT/no HT are columns.
```{r}
C = matrix(c(sum(cardiac$hxofHT == 0 & cardiac$gender == 0),
             sum(cardiac$hxofHT == 0 & cardiac$gender == 1),
             sum(cardiac$hxofHT == 1 & cardiac$gender == 0),
             sum(cardiac$hxofHT == 1 & cardiac$gender == 1)),
           2, 2)

C
```

Let's run the Fisher exact test.
```{r}
fisher.test(C, alternative = "greater")
```

The $p$ value is very high, so we cannot reject the null hypothesis. In fact, if we had used the opposite hypothesis, that women have higher prevalence of hypertension, then we _could_ have rejected the null hypothesis.


## Problem 2

```{r}
maleBP = cardiac$basebp[cardiac$gender == 0]
femaleBP = cardiac$basebp[cardiac$gender == 1]
```

## (a)

Null Hypothesis $H_0$: women's mean BP = men's mean BP

Alt Hypothesis $H_1$: women's mean BP < men's mean BP

## (b)

```{r}
boxplot(maleBP, femaleBP, main = "Blood Pressure for Men vs. Women", names = c("Men", "Women"))
```

## (c)

```{r}
## Sample sizes of the two groups:
n = length(maleBP)
m = length(femaleBP)

## Calculate pooled standard deviation
sp = sqrt(((n-1) * var(femaleBP) + (m-1) * var(maleBP)) / (m + n - 2) * (1/m + 1/n))
```

The $t$ statistic is:
```{r}
(tStat = (mean(femaleBP) - mean(maleBP)) / sp)
```

This is a one-sided test on the left (a.k.a. "less than"). The $p$ value is:
```{r}
(pValue = pt(tStat, df = m + n - 2))
```

We cannot reject the null hypothesis ($p > 0.05$). In fact, if we had tested
the opposite hypothesis (men's BP < women's BP), then we would be able to
reject the null hypothesis. This indicates that we had the hypothesis wrong.


## (d)

```{r}
eventBP = cardiac$basebp[cardiac$any.event == 0]
noEventBP = cardiac$basebp[cardiac$any.event == 1]
```

Null Hypothesis $H_0$: event mean BP = no-event mean BP

Alt Hypothesis $H_1$: event mean BP > no-event mean BP

```{r}
boxplot(eventBP, noEventBP, main = "Blood Pressure for Event vs. No Event", names = c("Event", "No Event"))
```

```{r}
## Sample sizes of the two groups
n = length(eventBP)
m = length(noEventBP)

## Calculate pooled standard deviation
sp = sqrt(((n-1) * var(eventBP) + (m-1) * var(noEventBP)) / (m + n - 2) * (1/m + 1/n))
```

The $t$ statistic is:
```{r}
(tStat = (mean(eventBP) - mean(noEventBP)) / sp)
```

This is a one-sided test on the right (a.k.a. "greater than"). The $p$ value is:
```{r}
(pValue = 1 - pt(tStat, df = m + n - 2))
```

Again, we cannot reject the null hypothesis. (This time the "less than"
hypothesis would have also been non-significant.)


## (e)

```{r}
eventEF = cardiac$baseEF[cardiac$any.event == 0]
noEventEF = cardiac$baseEF[cardiac$any.event == 1]
```

Null Hypothesis $H_0$: event mean EF = no-event mean EF

Alt Hypothesis $H_1$: event mean EF < no-event mean EF

```{r}
boxplot(eventEF, noEventEF, main = "Ejection Fraction for Event vs. No Event", names = c("Event", "No Event"))
```

```{r}
## Sample sizes of the two groups
n = length(eventEF)
m = length(noEventEF)

## Calculate pooled standard deviation
sp = sqrt(((n-1) * var(eventEF) + (m-1) * var(noEventEF)) / (m + n - 2) * (1/m + 1/n))
```

The $t$ statistic is:
```{r}
(tStat = (mean(eventEF) - mean(noEventEF)) / sp)
```


This is a one-sided test on the left (a.k.a. "less than"). The $p$ value is:
```{r}
(pValue = pt(tStat, df = m + n - 2))
```

On this one, we can reject the null hypothesis (p < 0.05). So, the data
supports that people who have a cardiac event have less efficient heart
pumping.


## Problem 3

## (a) 

$H_1 : \rho(X, Y) > 0$

Alternative hypothesis is that there is positive correlation.

## (b)

Scatter plot:

```{r}
n = dim(iris)[1]
plot(iris$Petal.Length, iris$Sepal.Length, xlab = "Petal Length", ylab = "Sepal Length", main = "Iris Sepal vs. Petal Length", lwd = 2)
```

## (c)

Critical value is $t_{1 - \alpha} = F^{-1}(1 - 0.05)$, where $F^{-1} is quantile function of $t(n-2)$ distribution.

```{r}
(criticalValue = qt(1 - 0.05, df = n - 2))
```

## (d) 

Use the formula to convert correlation into a $t$ statistic:

```{r}
r = cor(iris$Petal.Length, iris$Sepal.Length)
(tStat = r * sqrt((n - 2) / (1 - r^2)))
```

## (e)

$p$ value is a  right-sided test:

```{r}
(pValue = 1 - pt(tStat, df = n - 2))
```

## (f) 

$H1 : \rho(X, Y) < 0$

Alternative is there is a negative correlation.

## (g)

Same steps as (b) - (d) above, just switch to left-sided test

```{r}
plot(iris$Sepal.Width, iris$Sepal.Length, xlab = "Sepal Width", ylab = "Sepal Length", main = "Iris Sepal Width vs. Length", lwd = 2)

(criticalValue = qt(0.05, df = n - 2))

r = cor(iris$Sepal.Width, iris$Sepal.Length)
(tStat = r * sqrt((n - 2) / (1 - r^2)))

(pValue = pt(tStat, df = n - 2))
```

## Problem 4

## (a)

```{r}
my.regression = function(x, y)
{
  n = length(x)
  beta.hat = (n * sum(x * y) - sum(x)*sum(y)) / (n * sum(x^2) - sum(x)^2)

  ## alternative 1:
  ## beta.hat = cov(x, y) / var(x)

  ## alternative 2:
  ## beta.hat = cor(x, y) * sd(y) / sd(x)

  alpha.hat = mean(y) - beta.hat * mean(x)

  list(alpha.hat = alpha.hat, beta.hat = beta.hat)
}
```

```{r}
model = my.regression(faithful$waiting, faithful$eruptions)
plot(faithful$waiting, faithful$eruptions, xlab = "Waiting Time", ylab = "Eruption Time", main = "Old Faithful Data", lwd = 2)
abline(model$alpha.hat, model$beta.hat, col = 'red', lwd = 3)
```

## (b) 

Prediction based on the linear model: y = alpha + beta * x
## 10 pts
(predY = model$alpha.hat + model$beta.hat * 82)

## (c)

Note: they match!
```{r}
m = lm(faithful$eruptions ~ faithful$waiting)
print(m$coefficients)
print(model)
```

## (d) 

We can conclude that the slope is significantly different from zero, so there is a relationship between the waiting time and the eruption time.
```{r}
summary(m)
```