---
title: "R Markdown Example"
author: "Tom Fletcher"
date: "January 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
## This first bit is just setting up options for R.
## You can include this chunk in your documents as is.
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

The syntax below is how we insert a "chunk" of R code. The R code will appear formatted in the final document in a box, and if it produces text or a figure as output, that will appear after it in the document.

```{r}
summary(faithful)
```

Here is a plot example. Notice that `echo = FALSE` will prevent the code block from being printed in the document, but the output of the code (in this case a figure) will still appear.

```{r, echo=FALSE}
plot(faithful, main = "Old Faithful Data")
```

We can also include the result of an R command inside a line of text like this: The correlation between waiting time and eruption time in the Old Faithful data is `r cor(faithful$waiting, faithful$eruptions)`.

Here is the output of a more involved command, linear regression on the Old Faithful data.

```{r}
model = lm(waiting ~ eruptions, data = faithful)
summary(model)
```

Now we modify the plot above to include the regression line.

```{r, echo=FALSE}
plot(faithful, main = "Old Faithful Data")
abline(model, col = 'red', lwd = 3)
```

We can also insert mathematical notation into our document using LaTeX like so:
$$F(a) = \int_{-\infty}^a f(x) dx.$$

You can also write inline mathematics: $e^{i \pi} = -1$.